---
title: "Two sample proportions"

---


<!-- libraries -->
```{r,include=FALSE,purl=FALSE}
library(knitr) # for include_graphics() 
library(dplyr)
library(forcats)
library(readr)
library(ggplot2)
library(broom) 
library(tidyr)
library(ggplot2)
library(tibble)
```


```{r, echo=F, fig.align='center', out.width="50%"}
# students, dont worry about this code
library(ggplot2)
ggplot(data = data.frame(lower = -0.6, upper = 14.4, estimate = 11-4, group = "plus 4"), 
       aes(x = group, y = estimate)) +
  geom_point() +
  geom_segment(aes(xend = group, y = lower, yend = upper)) +
  geom_hline(aes(yintercept = 0), lty = 2) +
  labs(y = "Difference between vaccine and placebo", x = "") +
  theme_minimal(base_size = 15)
```

### Example using prop.test()

|         | Flu | no Flu | Total | $\hat{p}$ |
|---------|:---:|:------:|-------|-----------|
| Vaccine | 4   | 96     | 100   | 0.04      |
| Placebo | 11  | 89     | 100   | 0.11      |


```{r}
Flu  <- c( 11, 4) 
patients <- c( 100, 100)
prop.test(Flu, patients) ##first argument is the number of successes, the second is the number of observations
```

## Hypothesis testing - two samples binary data

### Hypothesis testing when you have two samples and binary data

$H_0: p_1 = p_2$ or $p_1-p_2=0$

$H_a:$

- $p_1 \neq p_2$ or $p_1-p_2\neq 0$ (two-sided)
- $p_1 > p_2$ or $p_1-p_2 > 0$ (one sided upper tail)
- $p_1 < p_2$ or $p_1-p_2 < 0$ (one sided lower tail)

### What does it mean to assume the null is true?

- If the null hypothesis is true, then $p_1$ is truly equal to $p_2$. In this
case, our best estimate of the underlying proportion that they are both equal to
is 

$$\hat{p} = \frac{\text{no. successes in both samples}}{\text{no. individuals in both samples}}$$

- Also, our best guess at the SE for $\hat{p}$ is:

$\sqrt{\frac{\hat{p}(1-\hat{p})}{n_1} + \frac{\hat{p}(1-\hat{p})}{n_2}}$

$\sqrt{\hat{p}(1-\hat{p})\big(\frac{1}{n_1}+\frac{1}{n_2})}$

This is the formula for the SE for the difference between two proportions but 
we have substituted $\hat{p}$ for $p_1$ and $p_2$.

### Hypothesis testing when you have two samples and binary data

Using the information from the previous slide, we can create the z-test for the
difference between two proportions as:

$$z = \frac{\hat{p_1} - \hat{p_2}}{\sqrt{\hat{p}(1-\hat{p})\Big(\frac{1}{n_1} + \frac{1}{n_2}\Big)}}$$

Use this test when the counts of successes and failures are $\geq$ 5 in both 
samples

### Example of hypothesis testing when you have two samples and binary data

Recall the RCT data on the occurrence of DVT between Fragamin vs. placebo groups:

|         | DVT | no DVT | Total | $\hat{p}$ |
|---------|-----|--------|-------|-----------|
| Fragamin| 42  | 1476   | 1518  | 0.0277    |
| Placebo | 73  | 1400   | 1473  | 0.0496    |

$H_0: p_1 = p_2$, or that the rate of DVT is the same between Fragamin and placebo
groups.

Suppose you're interested in knowing whether these two groups had different rates
of DVT. Then, $H_a: p_1 \neq p_2$

### Example of hypothesis testing when you have two samples and binary data

1. Compute $\hat{p} = \frac{42 + 73}{1518+1473} = \frac{115}{2991} = 0.03844868$
2. Compute the SE: $\sqrt{0.0384(1-0.0384)\big(\frac{1}{1518}+\frac{1}{1473}\big)} = 0.007032308$
3. Compute the test statistic: 
$$z = \frac{ 0.04955872 - 0.02766798}{0.007032308} = 3.11$$
4. Calculate the p-value
```{r}
pnorm(q = 3.112881, lower.tail = F)*2
```

## Two sample hypothesis testing in R

### Example of hypothesis testing when you have two samples and binary data

```{r}
prop.test(x = c(42, 73), # x is a vector of number of successes
          n = c(1518, 1473)) # n is a vector of sample sizes
```

### Example of hypothesis testing when you have two samples and binary data

- R gives a slightly different p-value because it has a continuity correction.
- This is okay. If you want to use R to check your hand calculation, you need
to add the argument `correct = F` to the calculation.

