---
title: " The chi-squared goodness of fit"

---

<!-- libraries -->
```{r,include=FALSE,purl=FALSE}
library(knitr) # for include_graphics() 
library(dplyr)
library(forcats)
library(readr)
library(ggplot2)
library(broom) 
library(tidyr)
library(ggplot2)
library(tibble)
```


### Jury Selection example

Suppose that the following number of people were selected for jury duty in the 
previous year, in a county where jury selection was supposed to be random. 

| Ethnicity      | White | Black | Latinx     | Asian      | Other | Total |
|-----------------|-------|-------|------------|------------|-------|-------|
| Number selected | 1920  | 347   | 19         | 84         | 130   | 2500  |

You read online about concerns that jury was not selected randomly. How can
you test this evidence?

- Example derived from this [video](https://www.youtube.com/watch?v=Uk36WGxujkc).

### Jury Selection example

Consider the distribution of race/ethnicity in the county overall:

| Ethnicity          | White | Black | Latinx | Asian  | Other | Total |
|---------------------|-------|-------|--------|--------|-------|-------|
| % in the population | 42.2% | 10.3%| 25.1%  | 17.1%   | 5.3% | 100%  |



How do we determine the counts that are **expected** (**E**) under the assumption that selection was random?:

| Ethnicity      | White | Black | Latinx     | Asian      | Other | Total |
|-----------------|-------|-------|------------|------------|-------|-------|
| Expected count  |       |       |            |            |       | 2500  |


### Jury Selection example

| Ethnicity          | White | Black | Latinx | Asian  | Other | Total |
|---------------------|-------|-------|--------|--------|-------|-------|
| % in the population | 42.2% | 10.3%| 25.1%  | 17.1%   | 5.3% | 100%  |

- To fill in the table, multiple the total size of the jury by the % of the 
population of each race/ethnicity:

Expected counts under the assumption that selection is random from the county:

| Ethnicity      | White       | Black     | Latinx     | Asian      | Other       | Total |
|:---------------:|:-----------:|:-----------:|:----------:|:----------:|:-----------:|:-----:|
| Expected count  | 2500 $\times$ 0.422 | 2500 $\times$ 0.103 | 2500 $\times$ 0.251 | 2500 $\times$ 0.171| 2500 $\times$ 0.053 | 2500  |
| = | 1055  | 257.5 | 627.5      |  427.5     | 132.5 | 2500  |


### Jury Selection example

How far off does the **observed counts** of race/ethnicities in the sample differ
from what we would expect if the jury had been selected randomly?

### Jury Selection example

Here are the counts we **observed** (**O**):

| Ethnicity      | White | Black | Latinx     | Asian      | Other | Total |
|-----------------|-------|-------|------------|------------|-------|-------|
| Observed count  | 1920  | 347   | 19         | 84         | 130   | 2500  |

Which we can compare to our **expected** (**E**):

| Ethnicity      | White       | Black     | Latinx     | Asian      | Other       | Total |
|:---------------:|:-----------:|:-----------:|:----------:|:----------:|:-----------:|:-----:|
| Expected count  |  1055  | 257.5 | 627.5      |  427.5     | 132.5 | 2500  |



### Jury Selection example

This plot shows the deviations between the observed and expected number of jurors.
What is the chance of observed deviations of these magnitudes (or larger) under the
null  hypothesis?

```{r, out.width="80%", fig.align='center', echo=F}

library(ggplot2)
observed_counts <- c(1920, 347,   19,    84, 130)
expected_counts <- c(1055, 257.5, 627.5, 427.5, 132.5)
race <- c("White", "Black", "Latinx", "Asian", "Other")
jury_data <- data.frame("observed_counts" = observed_counts, "expected_counts" = expected_counts, "race" = race)

ggplot(jury_data, aes(y = race, x = observed_counts)) + 
  geom_segment(aes(xend = expected_counts, yend = race)) +
  geom_point(aes(col = "Observed")) + 
  geom_point(aes(col = "Expected", x = expected_counts)) +
  theme_minimal(base_size = 15) +
  labs(x = "Observed (teal) vs. expected (pink) number of jurors",
       title = "Deviations between observed and expected number of jurors")

```

### The Chi-square distribution

The chi-square distribution is a new distribution to us. Like the t-distribution,
the chi-square distribution has a degrees of freedom parameter that changes the shape of the distribution. The degrees of freedom is equal to the number of groups (here, race/ethnicities) - 1.
Or, $df=k-1$.

```{r, out.width="60%", fig.align='center', echo=F}
# ggplot of the distribution at varying degrees of freedom

ggplot(data = data.frame(x = c(0.1, 30)), aes(x)) + 
  #geom_area(stat = "function", fun = dchisq, args = list(df=9),
      #      xlim = c(12, 20), fill = "grey") +
  stat_function(fun = dchisq, n = 101, args = list(df=2), aes(col = "df=2")) +
  stat_function(fun = dchisq, n = 101, args = list(df=4), aes(col = "df=4")) +
  stat_function(fun = dchisq, n = 101, args = list(df=9), aes(col = "df=9")) +
  stat_function(fun = dchisq, n = 101, args = list(df=15), aes(col = "df=15")) +
  theme_minimal(base_size = 15) + 
  theme(aspect.ratio=1) +
  labs(y = "Density", title = "Chi-squared distributions")
```

### The shape of the Chi-square
- As the `df` is increased, the distribution's central tendency moves to the right.
- This means that there will be more probability out in the right tail when the
degrees of freedom is higher. 
- The chi-square distribution is also positive. We only ever compute upper tail 
probabilities for the chi-square test because there is only one form to the $H_a$.

### Back to the jury example

**State the null and alternative hypotheses.** 

- The null hypothesis is that the proportions of each race/ethnicity in the jury 
pool is the same as the proportion of each group in the county. That is:

$H_0: p_{white} = 42.2\% \text{, }p_{black}=10.3\%\text{, }p_{latinx}=25.1\%\text{, }p_{asian}=17.1\%\text{, }p_{other}=5.3\%$

$H_a:$ At least one of $p_k$ is different than specified in $H_0$, for k being 
one of white, black, latinx, asian, or other.

### Back to the jury example

**Calculate the chi-square statistic using the jury data.**

| Ethnicity      | White | Black | Latinx     | Asian      | Other | Total |
|-----------------|-------|-------|------------|------------|-------|-------|
| O               | 1920  | 347   | 19         | 84         | 130   | 2500  |
| E               | 1055  | 257.5 | 627.5      | 427.5      | 132.5 | 2500  |

$\chi^2 = \sum_{i=1}^k\frac{(O_i-E_i)^2}{E_i}$

$\chi^2 = \frac{(1920-1055)}{1055}^2 + \frac{(347-257.5)}{257.5}^2 + \frac{(19-627.5)}{627.5}^2 + \frac{(84-427.5)}{427.5}^2 + \frac{(130-132.5)}{132.5}^2$

$\chi^2 = 709.218 + 31.10777 + 590.0753 + 276.0053 + 0.04716981$

$\chi^2 = 1606.454$

### Back to the jury example

**Calculate the p-value (what is the approprate degrees of freedom?).**

```{r}
pchisq(q = 1606.454, df = 4,lower.tail = F)
```

The probability of seeing this pool of people chosen for jury duty under the 
null hypothesis of random sampling from the county is so small that R rounded 
the p-value to 0! 

### Chi-square test in R

**Run the chi-square test using the `chisq.test` command in R.**

```{r}
chisq.test(x = c(1920, 347, 19, 84, 130), # x is vector of observed counts
           p = c(.422, .103, .251, .171, .053)) # p is probability under the null
```

### Interpretation

- Which race/ethnicities appear to deviate the most from what was expected under
the null hypothesis?
    - Compare the proportion observed vs. proportion expected
    - Compare the count observed vs. the count expected
    - Compare the 5 contributions to the chi-square test from each race/ethnicity.
    We see that whites, Latinx, and Asians contribute the most to the $\chi^2$
    statistic. This agrees with what we saw in the data visualization in terms
    of the size of the gaps between observed and expected counts.

### graph
```{r, echo=FALSE, out.width="70%"}
ggplot(jury_data, aes(y = race, x = observed_counts)) + 
  geom_segment(aes(xend = expected_counts, yend = race)) +
  geom_point(aes(col = "Observed")) + 
  geom_point(aes(col = "Expected", x = expected_counts)) +
  theme_minimal(base_size = 15) +
  labs(x = "Observed (teal) vs. expected (pink) number of jurors",
       title = "Deviations between observed and expected number of jurors")

```

## Statistics is Everywhere

### Example 2:  Alameda Chi-square test in R

**Run the chi-square test using the `chisq.test` command in R.**

```{r}
chisq.test(x = c(209299,5024,2012,6617,77668,194647,45509), # x is vector of observed counts
           p = c(.306,.009,.011,.054,.187,.323,.11)) # p is probability under the null
```


### Example 3:  cheating at dice?
Suppose there is a game in which the objective is to roll sixes as possible using 3 die.
Over 100 rolls, one of the players seems to be winning quite often, we see the following 

| Number of 6s    | 0     | 1     | 2    | 3    | 
|:---------------:|:-----:|:-----:|:----:|:----:|
| Observed rolls  |  47   | 35    | 15   |  3   |

We suspect they are using a loaded die or cheating in some way.  

Are they cheating?  Or just lucky (within the bounds of chance)?

Example derived from [this](http://www.stat.yale.edu/Courses/1997-98/101/chigf.htm) site

### Example 3:  cheating at dice?
What would we expect?  
The rolls of dice should follow a binomial distribution (# of successes in # trials)


$$P(X=k)={n\choose k}p^k(1-p)^{n-k}$$
What is P here?  What is K?

### Example 3:  cheating at dice?
Remember dbinom?

'dbinom(#successes,size,probability of success)'

This function calculates the probability of observing `x` successes when $X \sim Binom(n,p)$

```{r binoms, echo=TRUE}
Expect_0<-dbinom(0,size=3,prob=0.166666667)
Expect_1<-dbinom(1,size=3,prob=0.166666667)
Expect_2<-dbinom(2,size=3,prob=0.166666667)
Expect_3<-dbinom(3,size=3,prob=0.166666667)
Expected<-c(Expect_0,Expect_1,Expect_2,Expect_3)
Expected
``` 

### Example 3:  cheating at dice?

| Number of 6s    | 0     | 1     | 2    | 3    | 
|:---------------:|:-----:|:-----:|:----:|:----:|
| Observed rolls  |  47   | 35    | 15   |  3   |
| Expected rolls  |  57.9 | 34.7  | 6.9  |  0.46|


### Exampe 3:  cheating at dice?
```{r}
chisq.test(x = c(47,35,15,3), # x is vector of observed counts
           p = Expected) # p is probability under the null
```





