---
title: "From Z to T"

---

<!-- libraries -->
```{r,include=FALSE,purl=FALSE}
library(knitr) # for include_graphics() 
library(dplyr)
```

### Introducing the t distribution
```{r, echo= F, fig.align='center', out.width="80%", warning=FALSE, message=FALSE}
#students do not need to know this code
library(ggplot2)
ggplot(data = data.frame(x = c(-3, 3)), aes(x)) +
  stat_function(fun = dnorm, n = 101, args = list(mean = 0, sd = 1), aes(col = "N(0,1)")) +
  theme_minimal(base_size = 15) +
  stat_function(fun = dt, n = 101, args = list(df = 2), aes(col = "t(df=2)")) +
  stat_function(fun = dt, n = 101, args = list(df = 4), aes(col = "t(df=4)")) +
  stat_function(fun = dt, n = 101, args = list(df = 9), aes(col = "t(df=9)")) +
  scale_color_manual(values = c("black", "#1b9e77", "#d95f02","blue")) +
  labs(y = "Density", title = "Comparing N(0,1) to t-distributions")
```

### Guess the R functions

```{r t-functions, eval=F, echo=TRUE}
pt(q = , df = , lower.tail = )
qt(p = , df = , lower.tail = )
```

Supposing we had $n=100$, what is $t^*$ for a 95% confidence interval?

```{r, eval=F, echo=TRUE}
qt(p = 0.975, df = 99)
```
### Example: Testosterone and obesity in adolescent males (pg 422 B&M Ed 4)

Here are the data for $n=25$ adolescent males between the ages of 14 and 20:

```{r, message=F, warning=F, echo=TRUE}
library(dplyr)
testosterone <- c(0.30, 0.24, 0.19, 0.17, 0.18, 0.23, 0.24, 0.06, 0.15,
                  0.17, 0.18, 0.17, 0.15, 0.12, 0.25, 0.25, 0.25, 0.32, 
                  0.35, 0.37, 0.39, 0.46, 0.49, 0.42, 0.36)
dat_test <- data.frame(testosterone)
```

### Example: Testosterone and obesity in adolescent males (pg 422 B&M Ed 4)

Use R to calculate a 95% confidence interval for testosterone. We can do this 
using `summarize`

```{r, echo=TRUE}
dat_test %>% summarize(sample_mean = mean(testosterone),
                       sample_sd = sd(testosterone),
                       sample_size = length(testosterone),
                       sample_se = sample_sd/sqrt(sample_size))
```


### Example: Testosterone and obesity in adolescent males (pg 422 B&M Ed 4)
We still need the $t^*$ value:

```{r, echo=TRUE}
t_star <- qt(p = 0.975, df = 24)
t_star
```

### Example: Testosterone and obesity in adolescent males (pg 422 B&M Ed 4)

Expand the previous code chunk to calculate the margin of error (which uses the
critical $t^*$ value), and then calculate the lower and upper CI

```{r, echo=TRUE}
dat_test %>% summarize(sample_mean = mean(testosterone),
                       sample_sd = sd(testosterone),
                       sample_size = length(testosterone),
                       sample_se = sample_sd/sqrt(sample_size),
                       margin_of_error = sample_se*t_star, 
                       lower_CI = sample_mean - margin_of_error, 
                       upper_CI = sample_mean + margin_of_error)
```

### Hypothesis testing with unknown $\sigma$ using the t-test
In terms of a variable T having the $t(n-1)$ distribution, the p-value for a
test of $H_0$ against

$H_a$: $\mu > \mu_0$ is $P(T \geq t)$

```{r, out.width="30%", echo=F}
upper <- ggplot(data = data.frame(x = c(-4, 4)), aes(x)) + 
  geom_area(stat = "function", fun = dt, args = list(df=10),
            xlim = c(1.7, 4), fill = "orange") +
  stat_function(fun = dt, n = 101, args = list(df=10)) +
  theme_minimal(base_size = 15) + 
  labs(title = "One-sided (above)", y = "Density", x = "") +
  theme(aspect.ratio=1) 
upper
```


### Hypothesis testing with unknown $\sigma$ using the t-test
$H_a$: $\mu < \mu_0$ is $P(T \leq t)$

```{r, out.width="30%", echo=F}
lower <- ggplot(data = data.frame(x = c(-4, 4)), aes(x)) + 
  geom_area(stat = "function", fun = dt, args = list(df=10),
            xlim = c(-4, -1.7), fill = "orange") +
  stat_function(fun = dt, n = 101, args = list(df=10)) +
  theme_minimal(base_size = 15) + 
  labs(title = "One-sided (below)", y = "Density", x = "") +
  theme(aspect.ratio=1)
lower
```

### Hypothesis testing with unknown $\sigma$ using the t-test
$H_a$: $\mu \neq \mu_0$ is $2\times P(T \geq |t|)$

```{r, out.width="30%", echo=F}
both <- ggplot(data = data.frame(x = c(-3, 3)), aes(x)) + 
  geom_area(stat = "function", fun = dt, args = list(df=25),
            xlim = c(1.7, 3), fill = "orange") +
  geom_area(stat = "function", fun = dt, args = list(df=25),
            xlim = c(-3, -1.7), fill = "orange") +
  stat_function(fun = dt, n = 101, args = list(df=25)) +
  theme_minimal(base_size = 15) + 
  labs(title = "Two-sided", y = "Density", x = "") +
  theme(aspect.ratio=1)

#library(patchwork)
both
#upper + lower + both + plot_layout() 
```

## Example and t-testing in R

### Example of a t-test (pg 426 B&M Ed 4)

Here are 18 measures of pulse wave velocity (PWV) from a sample of children
diagnosed with progeria, a genetic disorder that produces rapid aging.

```{r, echo=TRUE}
pwv <- c(18.8, 17.6, 17.5, 16.0, 14.8, 14.1, 13.7, 13.1, 12.9,
         12.9, 12.4, 10.1, 9.3,  9.1,  8.3,  8.3,  7.9,  7.2)

pwv_dat <- data.frame(pwv)
```


### Example of a t-test (pg 426 B&M Ed 4)

pwv measures greater than 6.6 are considered abnormally high. We would like to 
test the hypothesis that the mean for this subset of children is abnormally high.

That is: $H_0: \mu = 6.6$ and $H_a: \mu>6.6$

### Look at the data and see if there is evidence against the null hypothesis

```{r, fig.align='center', out.width="80%", echo=F}
ggplot(pwv_dat, aes(x = pwv)) +
  theme_minimal(base_size = 15) +
  geom_histogram(binwidth = 2, fill = "white", col = "black") +
  geom_vline(xintercept = 6.6, lty = 2, col = "blue") +
  geom_text(aes(x=6, y = 3.5, label = "H0"), check_overlap = T, col = "blue") +
  geom_vline(xintercept = 12.44, lty = 2) +
  geom_text(aes(x=11.8, y = 3.5, label = "sample\nmean"), check_overlap = T) 
```




### Example of a one-sided t-test (pg 426 B&M Ed 4)


```{r, echo=TRUE}
pwv_dat %>% 
  summarize(sample_mean = mean(pwv),
            sample_sd = sd(pwv),
            sample_size = length(pwv),
            sample_se = sample_sd/sqrt(sample_size),
            t_test = (sample_mean - 6.6)/sample_se, 
            p_value = 1 - pt(t_test, df = sample_size - 1))
```

### There's a function for that...

Rather than doing the test using summarize, we could have R do it for us using
`t.test`:

```{r, echo=TRUE}
t.test(x = pwv_dat %>% pull(pwv), alternative = "greater", mu = 6.6)
```
