---
title: "Power: Inference in Practice"
---


<!-- libraries -->
```{r,include=FALSE,purl=FALSE}
library(knitr) # for include_graphics() 
library(dplyr)
```


## Calculating power in R

### Example of calculating power

Suppose you have a known standard deviation $\sigma = 1$. $H_0: \mu = 0$ vs. 
$H_a: \mu > 0.8$ and choose $\alpha = 0.05$. Calculate the power when $n=10$.

You can calculate the minimum z-value required to reject $H_0$:

```{r}
qnorm(p = 0.05, mean = 0, sd = 1/sqrt(10), lower.tail = F)
```

So for any z-test with this value or higher, you will reject $H_0$ in favor of $H_a$.

This is often called $Z_{\alpha}$

### Example of calculating power

Now suppose that $H_a$ is true. The test will reject $H_0$ about what percent of the 
time when $H_a$ is true? To calculate this probability, we take the value from the 
previous calculation and calculate the *probability* above its value under $H_a$:

```{r}
pnorm(q = 0.5201484, mean = 0.8, 1/sqrt(10), lower.tail = F)
```

### Example of calculating power, illustrated

```{r, echo= F, fig.align='center', out.width="80%"}
library(ggplot2)
ggplot(data = data.frame(x = c(-1.3, 2)), aes(x)) + 
    geom_area(stat = "function", fun = dnorm, args = list(mean = 0, sd = 1/sqrt(10)), 
            fill = "red", xlim = c(0.5201484, 1.1)) +
  stat_function(fun = dnorm, n = 101, args = list(mean = 0, sd = 1/sqrt(10))) +
  labs(y = "density") + 
  theme_minimal(base_size = 15) +
  labs(title = "Calculating power") +
  geom_area(stat = "function", fun = dnorm, args = list(mean = 0.8, sd = 1/sqrt(10)), 
            fill = "blue", xlim = c(0.5201484, 2.0), alpha = 0.5) +
  stat_function(fun = dnorm, n = 101, args = list(mean = 0.8, sd = 1/sqrt(10))) +
  geom_text(aes(x = 0, y = 1.2), label = "H0", check_overlap = T) +
  geom_text(aes(x = 0.8, y = 1.2), label = "Ha", check_overlap = T)

```


## Size of the difference

### Effect of changing the $\mu_{1}$

Imagine we our $H_{0}$ is a standard normal (mean=0, SD=1) and we set our $\alpha$ at 0.05.  

If the true mean of our sampled population is 1.7 standard deviations above the $\mu_{0}$, 


what does our $\beta$ look like?

### Effect of changing the $\mu_{1}$
```{r powerpic, echo=FALSE}
##students you do not need to know this code

##graph power
x <- seq(-4, 4, length=1000)
hx <- dnorm(x, mean=0, sd=1)

plot(x, hx, type="n", xlim=c(-4, 8), ylim=c(0, 0.5), 
     ylab = "",
     xlab = "",
     main= expression(paste("Type II (", beta, ") error")), axes=FALSE)
axis(1, at = c(-qnorm(.025), 0, -4), 
     labels = expression("p-value", 0, -infinity ))

shift = qnorm(1-0.025, mean=0, sd=1)*1.7
xfit2 <- x + shift
yfit2 <- dnorm(xfit2, mean=shift, sd=1)

# Print null hypothesis area
col_null = "#AAAAAA"
polygon(c(min(x), x,max(x)), c(0,hx,0), col=col_null, lwd=2, density=c(10, 40), angle=-45, border=0)
lines(x, hx, lwd=2, lty="dashed", col=col_null)


# The alternative hypothesis area

## The red - underpowered area
lb <- min(xfit2)
ub <- round(qnorm(.975),2)
col1 = "#CC2222"

i <- xfit2 >= lb & xfit2 <= ub
polygon(c(lb,xfit2[i],ub), c(0,yfit2[i],0), col=col1)

## The green area where the power is
col2 = "#22CC22"
i <- xfit2 >= ub
polygon(c(ub,xfit2[i],max(xfit2)), c(0,yfit2[i],0), col=col2)

# Outline the alternative hypothesis
lines(xfit2, yfit2, lwd=2)

axis(1, at = (c(ub, max(xfit2))), labels=c("", expression(infinity)), 
     col=col2, lwd=1, lwd.tick=FALSE)

legend("topright", inset=.015, title="Color",
       c("Null hypothesis","Type II error", "True"), fill=c(col_null, col1, col2), 
       angle=-45,
       density=c(20, 1000, 1000), horiz=FALSE)

abline(v=ub, lwd=2, col="#000088", lty="dashed")

arrows(ub, 0.45, ub+1, 0.45, lwd=3, col="#008800")
arrows(ub, 0.45, ub-1, 0.45, lwd=3, col="#880000")
```

### Effect of changing the $\mu_{1}$ 

What happens if the "true" mean is closer to the Null?

### Effect of changing the $\mu_{1}$

```{r powerpic2, echo=FALSE}


##graph power
x <- seq(-4, 4, length=1000)
hx <- dnorm(x, mean=0, sd=1)

plot(x, hx, type="n", xlim=c(-4, 8), ylim=c(0, 0.5), 
     ylab = "",
     xlab = "",
     main= expression(paste("Type II (", beta, ") error")), axes=FALSE)
axis(1, at = c(-qnorm(.025), 0, -4), 
     labels = expression("p-value", 0, -infinity ))

shift = qnorm(1-0.025, mean=0, sd=1)*1.2
xfit2 <- x + shift
yfit2 <- dnorm(xfit2, mean=shift, sd=1)

# Print null hypothesis area
col_null = "#AAAAAA"
polygon(c(min(x), x,max(x)), c(0,hx,0), col=col_null, lwd=2, density=c(10, 40), angle=-45, border=0)
lines(x, hx, lwd=2, lty="dashed", col=col_null)


# The alternative hypothesis area

## The red - underpowered area
lb <- min(xfit2)
ub <- round(qnorm(.975),2)
col1 = "#CC2222"

i <- xfit2 >= lb & xfit2 <= ub
polygon(c(lb,xfit2[i],ub), c(0,yfit2[i],0), col=col1)

## The green area where the power is
col2 = "#22CC22"
i <- xfit2 >= ub
polygon(c(ub,xfit2[i],max(xfit2)), c(0,yfit2[i],0), col=col2)

# Outline the alternative hypothesis
lines(xfit2, yfit2, lwd=2)

axis(1, at = (c(ub, max(xfit2))), labels=c("", expression(infinity)), 
     col=col2, lwd=1, lwd.tick=FALSE)

legend("topright", inset=.015, title="Color",
       c("Null hypothesis","Type II error", "True"), fill=c(col_null, col1, col2), 
       angle=-45,
       density=c(20, 1000, 1000), horiz=FALSE)

abline(v=ub, lwd=2, col="#000088", lty="dashed")

arrows(ub, 0.45, ub+1, 0.45, lwd=3, col="#008800")
arrows(ub, 0.45, ub-1, 0.45, lwd=3, col="#880000")
```

### Effect of changing the $\mu_{1}$

What happens if the "true" mean is further from the null?

### Effect of changing the $\mu_{1}$

```{r powerpic3, echo=FALSE}

##graph power
x <- seq(-4, 4, length=1000)
hx <- dnorm(x, mean=0, sd=1)

plot(x, hx, type="n", xlim=c(-4, 8), ylim=c(0, 0.5), 
     ylab = "",
     xlab = "",
     main= expression(paste("Type II (", beta, ") error")), axes=FALSE)
axis(1, at = c(-qnorm(.025), 0, -4), 
     labels = expression("p-value", 0, -infinity ))

shift = qnorm(1-0.025, mean=0, sd=1)*2.4
xfit2 <- x + shift
yfit2 <- dnorm(xfit2, mean=shift, sd=1)

# Print null hypothesis area
col_null = "#AAAAAA"
polygon(c(min(x), x,max(x)), c(0,hx,0), col=col_null, lwd=2, density=c(10, 40), angle=-45, border=0)
lines(x, hx, lwd=2, lty="dashed", col=col_null)


# The alternative hypothesis area

## The red - underpowered area
lb <- min(xfit2)
ub <- round(qnorm(.975),2)
col1 = "#CC2222"

i <- xfit2 >= lb & xfit2 <= ub
polygon(c(lb,xfit2[i],ub), c(0,yfit2[i],0), col=col1)

## The green area where the power is
col2 = "#22CC22"
i <- xfit2 >= ub
polygon(c(ub,xfit2[i],max(xfit2)), c(0,yfit2[i],0), col=col2)

# Outline the alternative hypothesis
lines(xfit2, yfit2, lwd=2)

axis(1, at = (c(ub, max(xfit2))), labels=c("", expression(infinity)), 
     col=col2, lwd=1, lwd.tick=FALSE)

legend("topright", inset=.015, title="Color",
       c("Null hypothesis","Type II error", "True"), fill=c(col_null, col1, col2), 
       angle=-45,
       density=c(20, 1000, 1000), horiz=FALSE)

abline(v=ub, lwd=2, col="#000088", lty="dashed")

arrows(ub, 0.45, ub+1, 0.45, lwd=3, col="#008800")
arrows(ub, 0.45, ub-1, 0.45, lwd=3, col="#880000")
```

## Example From Pagano and Gauvreau "Principles of Biostatistics"

### Example: from Pagano text ch. 10
Suppose we have a sample of 25 20 to 74 year old males from the United States.  We know that 20-24 year old males have a mean serum cholesterol level of 180 mg/100ml.  Since cholesterol tends to increase with age, we would expect the mean cholesterol in our sample (of 20-74 year olds) to be higher than 180 mg/100ml.  We also assume the standard deviation in the population is 46 mg/100ml.  

This is a one-sided hypothesis with $\alpha$ = 0.05, we can use r to get the critical value in Z 
```{r setcon, }
qnorm(0.95, 0,1)
```

so $H_{0}$ would be rejected at $Z\ge 1.645$

At what mean value of cholesterol would we decide to reject the null?

### Example: from Pagano text ch. 10

$$Z=\frac{\overline{x}-\mu_0}{\frac{\sigma}{\sqrt{n}}}$$

$$1.645=\frac{\overline{x}-180}{\frac{46}{\sqrt{25}}}$$
Solve this for $\overline{x}$


$$\bar{x} = 1.645 \times\frac{46}{\sqrt{25}}+180=195.1$$

### Example: from Pagano text ch. 10
So here we have our null distribution with the value at which we reject the null
```{r powerpic4, echo=FALSE, out.width="90%"}

##graph power
x <- seq(-4, 4, length=1000)
hx <- dnorm(x, mean=0, sd=1)

plot(x, hx, type="n", xlim=c(-4, 8), ylim=c(0, 0.5), 
     ylab = "",
     xlab = "",
     main= expression(paste("Critical value, 195.1")), axes=FALSE)
axis(1, at = c(-qnorm(.025), 0, -4), 
     labels = expression("p-value", 180, -infinity ))

shift = qnorm(1-0.025, mean=0, sd=1)*2.4
xfit2 <- x + shift
yfit2 <- dnorm(xfit2, mean=shift, sd=1)

# Print null hypothesis area
col_null = "#AAAAAA"
polygon(c(min(x), x,max(x)), c(0,hx,0), col=col_null, lwd=2, density=c(10, 40), angle=-45, border=0)
lines(x, hx, lwd=2, lty="dashed", col=col_null)


abline(v=1.645, lwd=2, col="#000088", lty="dashed")

```

What is our $\beta$ ? 

### Example: from Pagano text ch. 10

We must choose a value at which to evaluate $\beta$.  Here we will choose an alternate hypothesis that $\mu_{1}=211$.  Since we know a sample mean less than 195.1 causes us to fail to reject $H_{0}$ we need to calculate the proportion of a distribution centered at 211mg/100ml that would be below this value.

$$Z=\frac{195.1-211}{\frac{46}{\sqrt{25}}}$$

$$Z=-1.73$$


### Example: from Pagano text ch. 10
Using R to calculate the probability, 
```{r exnorm, echo=TRUE}
pnorm(-1.73, 0,1)
```

Thus $\beta$ P(do not reject null(180)|Null is false (true population mean is 211)) is ~0.042

Remember that Power is 1-$\beta$ = P(reject null | null is false)

In this example, Power is 1-0.042 = 0.958

### Example: from Pagano text ch. 10

```{r powerpic5, echo=FALSE}

##graph power
x <- seq(-4, 4, length=1000)
hx <- dnorm(x, mean=0, sd=1)

plot(x, hx, type="n", xlim=c(-4, 8), ylim=c(0, 0.5), 
     ylab = "",
     xlab = "",
     main= expression(paste("Critical value, 195.1")), axes=FALSE)
axis(1, at = c(-qnorm(.025), 0, -4), 
     labels = expression("p-value", 180, -infinity ))

shift = qnorm(1-0.025, mean=0, sd=1)*2
xfit2 <- x + shift
yfit2 <- dnorm(xfit2, mean=shift, sd=1)

# Print null hypothesis area
col_null = "#AAAAAA"
polygon(c(min(x), x,max(x)), c(0,hx,0), col=col_null, lwd=2, density=c(10, 40), angle=-45, border=0)
lines(x, hx, lwd=2, lty="dashed", col=col_null)


# The alternative hypothesis area

## The red - underpowered area
lb <- min(xfit2)
ub <- round(qnorm(.95),2)
col1 = "#CC2222"

i <- xfit2 >= lb & xfit2 <= ub
polygon(c(lb,xfit2[i],ub), c(0,yfit2[i],0), col=col1)

## The green area where the power is
col2 = "#22CC22"
i <- xfit2 >= ub
polygon(c(ub,xfit2[i],max(xfit2)), c(0,yfit2[i],0), col=col2)

# Outline the alternative hypothesis
lines(xfit2, yfit2, lwd=2)

axis(1, at = (c(ub, max(xfit2))), labels=c("", expression(infinity)), 
     col=col2, lwd=1, lwd.tick=FALSE)

legend("topright", inset=.015, title="Color",
       c("Null hypothesis","Type II error", "True"), fill=c(col_null, col1, col2), 
       angle=-45,
       density=c(20, 1000, 1000), horiz=FALSE)

abline(v=ub, lwd=2, col="#000088", lty="dashed")

arrows(ub, 0.45, ub+1, 0.45, lwd=3, col="#008800")
arrows(ub, 0.45, ub-1, 0.45, lwd=3, col="#880000")
```

### Size of the difference
So we have we learned about power and $\beta$ ?

If we increase the size of the difference we are looking for, what happens to power?

What happens to $\beta$ as we increase the size of the difference?

## Size of the sample


### Remember what happens to our test statistic as sample size increases


$$Z=\frac{\overline{x}-\mu_0}{\frac{\sigma}{\sqrt{n}}}$$



### Increasing sample size
If we look at this graph from earlier in the slides, what do we expect to change if we increase our sample size from 25 to 100?
```{r size1, echo=FALSE, out.width="80%"}
##students you do not need to know this code

##graph power
x <- seq(-4, 4, length=1000)
hx <- dnorm(x, mean=0, sd=1)

plot(x, hx, type="n", xlim=c(-4, 8), ylim=c(0, 0.5), 
     ylab = "",
     xlab = "",
     main= expression(paste("Type II (", beta, ") error")), axes=FALSE)
axis(1, at = c(-qnorm(.025), 0, -4), 
     labels = expression("p-value", 0, -infinity ))

shift = qnorm(1-0.025, mean=0, sd=1)*1.7
xfit2 <- x + shift
yfit2 <- dnorm(xfit2, mean=shift, sd=1)

# Print null hypothesis area
col_null = "#AAAAAA"
polygon(c(min(x), x,max(x)), c(0,hx,0), col=col_null, lwd=2, density=c(10, 40), angle=-45, border=0)
lines(x, hx, lwd=2, lty="dashed", col=col_null)


# The alternative hypothesis area

## The red - underpowered area
lb <- min(xfit2)
ub <- round(qnorm(.975),2)
col1 = "#CC2222"

i <- xfit2 >= lb & xfit2 <= ub
polygon(c(lb,xfit2[i],ub), c(0,yfit2[i],0), col=col1)

## The green area where the power is
col2 = "#22CC22"
i <- xfit2 >= ub
polygon(c(ub,xfit2[i],max(xfit2)), c(0,yfit2[i],0), col=col2)

# Outline the alternative hypothesis
lines(xfit2, yfit2, lwd=2)

axis(1, at = (c(ub, max(xfit2))), labels=c("", expression(infinity)), 
     col=col2, lwd=1, lwd.tick=FALSE)

legend("topright", inset=.015, title="Color",
       c("Null hypothesis","Type II error", "True"), fill=c(col_null, col1, col2), 
       angle=-45,
       density=c(20, 1000, 1000), horiz=FALSE)

abline(v=ub, lwd=2, col="#000088", lty="dashed")

arrows(ub, 0.45, ub+1, 0.45, lwd=3, col="#008800")
arrows(ub, 0.45, ub-1, 0.45, lwd=3, col="#880000")
```

### Increasing sample size
```{r size2, echo=FALSE}
##students you do not need to know this code

##graph power
x <- seq(-4, 4, length=1000)
hx <- dnorm(x, mean=0, sd=.5)

plot(x, hx, type="n", xlim=c(-4, 8), ylim=c(0, 0.8), 
     ylab = "",
     xlab = "",
     main= expression(paste("Type II (", beta, ") error")), axes=FALSE)
axis(1, at = c(-qnorm(.025,0,.5), 0, -4), 
     labels = expression("p-value", 0, -infinity ))

shift = qnorm(1-0.025, mean=0, sd=1)*1.7
xfit2 <- x + shift
yfit2 <- dnorm(xfit2, mean=shift, sd=.5)

# Print null hypothesis area
col_null = "#AAAAAA"
polygon(c(min(x), x,max(x)), c(0,hx,0), col=col_null, lwd=2, density=c(10, 40), angle=-45, border=0)
lines(x, hx, lwd=2, lty="dashed", col=col_null)


# The alternative hypothesis area

## The red - underpowered area
lb <- min(xfit2)
ub <- round(qnorm(.975,0,.5),2)
col1 = "#CC2222"

i <- xfit2 >= lb & xfit2 <= ub
polygon(c(lb,xfit2[i],ub), c(0,yfit2[i],0), col=col1)

## The green area where the power is
col2 = "#22CC22"
i <- xfit2 >= ub
polygon(c(ub,xfit2[i],max(xfit2)), c(0,yfit2[i],0), col=col2)

# Outline the alternative hypothesis
lines(xfit2, yfit2, lwd=2)

axis(1, at = (c(ub, max(xfit2))), labels=c("", expression(infinity)), 
     col=col2, lwd=1, lwd.tick=FALSE)

legend("topright", inset=.015, title="Color",
       c("Null hypothesis","Type II error", "True"), fill=c(col_null, col1, col2), 
       angle=-45,
       density=c(20, 1000, 1000), horiz=FALSE)

abline(v=ub, lwd=2, col="#000088", lty="dashed")

arrows(ub, 0.45, ub+1, 0.45, lwd=3, col="#008800")
arrows(ub, 0.45, ub-1, 0.45, lwd=3, col="#880000")
```


## Calculating sample size


### Example 
For example, using our previous study of mean serum cholesterol levels, if we remember that we assumed the following:

$H_{0}$ :  $\mu\leq 180 mg/100ml$

$\alpha$:  0.01

$\sigma$: 46 

If the true population mean is as large as 211 and we want to risk only a 5% chance of failing to reject the null, so $\beta=0.05$ and power would be = $1-\beta=0.95$

### Calculating Sample Size
We start by finding the Z value at which we would reject $H_{0}$ at $\alpha$ = 0.01

We call this value $Z_{\alpha}$

```{r samp1, echo=TRUE}
qnorm(0.01, lower=FALSE)
```

Solve for $\overline{x}$

$$2.32 =\frac{\overline{x}-180}{\frac{46}{\sqrt{n}}}$$
$$\overline{x}=180+2.32(\frac{46}{\sqrt{n}})$$

### Calculating Sample Size

Next we find the Z value at which we would reject $H_{A}$ at $\beta$ = 0.05

We call this value $Z_{\beta}$

```{r samp2, echo=TRUE}
qnorm(0.05)
```
Solve for $\overline{x}$

$$-1.645 =\frac{\overline{x}-211}{\frac{46}{\sqrt{n}}}$$
$$\overline{x}=211-1.645(\frac{46}{\sqrt{n}})$$

### Calculating Sample Size

Because the value of $\overline{x}$ represents the same cutpoint for each of these, we can set the two equations equal to each other and solve for n.

$$180+2.32\left(\frac{46}{\sqrt{n}}\right)=211-1.645\left(\frac{46}{\sqrt{n}}\right)$$
$$\sqrt{n}(180-211)=(-2.32+(-1.645))*46$$
$$n=\left(\frac{(-2.32-1.645)*(46)}{(180-211)}\right)^2$$
$$n=34.6$$
As we cannot include 0.6 of a person, the convention is to round up.  So we would need 35 people in our sample.

### Calculating Sample Size

The more general way to express this sample size formula for a one sided Z test is

$$n=\left(\frac{(Z_{\alpha}+Z_{\beta})*(\sigma)}{(\mu_{1}-\mu_{0})}\right)^2$$
For a two sided Z test this would be 
$$n=\left(\frac{(Z_{(\frac{\alpha}{2})}+Z_{\beta})*(\sigma)}{(\mu_{1}-\mu_{0})}\right)^2$$


## Determining sample size for a margin of error

### Selecting an appropriate sample size 

How big does the sample size need to be to give a certain size of confidence interval?

Suppose you want your margin of error to equal $m$. What sample size do you need
to obtain a margin of error of $m$? 

You can re-frame the sample size formula for a two sided hypothesis test  

$$n=\left(\frac{(Z_{(\frac{\alpha}{2})}+Z_{\beta})*(\sigma)}{(\mu_{1}-\mu_{0})}\right)^2$$

to a margin of error

$$n=\Big(\frac{Z^*\sigma}{m}\Big)^2$$

### Example of calculation sample size

Body temperature has a known $\sigma=0.6$ degrees F. We want to estimate the mean
body temperature $\mu$ for healthy adults within $\pm0.05$ F with 95% confidence. 
How many healthy adults must we measure?

$$n=\Big(\frac{z^*\sigma}{m}\Big)^2$$
$$n=\Big(\frac{1.96\times 0.6}{0.05}\Big)^2 = 553.2$$

We must recruit 554 (round up!) healthy adults for this study.
