---
title: "Flavors of T"
---

<!-- libraries -->
```{r,include=FALSE,purl=FALSE}
library(knitr) # for include_graphics() 
library(dplyr)
library(readr)
```

### Example with heights from two different populations

Are the heights of US and Dutch-born men different?

```{r heights-men, fig.align='center', out.width="80%", warning=F, message=F, echo=F}
library(tidyverse)
set.seed(123)
dutch <- rnorm(n = 100, mean = 183, sd = 7.4)
usa <- rnorm(n = 100, mean = 175.5, sd = 7.4)

height_data_wide = data.frame(usa, dutch)
height_data <- data.frame(height = c(dutch, usa), country = c(rep("Dutch", 100), rep("USA", 100)))

height_data %>% group_by(country) %>% summarise(sample_mean = mean(height), 
                                                sample_sd = sd(height), 
                                                length = length(height))
```

### Example with heights from two different populations
```{r men2, fig.align='center', out.width="80%", warning=F, message=F, echo=F}
ggplot(height_data, aes(x = height)) + 
  geom_histogram(aes(fill = country), binwidth = 5, col = "black") +
  theme_minimal(base_size = 15) +
  facet_wrap(~country, nrow = 2)
```

### Example with heights from two different populations
If we take out the last line of that code, the histograms will be plotted on one grid.  Notice what happens to the columns...
```{r men3, fig.align='center', out.width="80%", warning=F, message=F, echo=F}
ggplot(height_data, aes(x = height)) + 
  geom_histogram(aes(fill = country), binwidth = 5, col = "black") +
  theme_minimal(base_size = 15)
```

### Example with heights from two different populations  
```{r men4, fig.align='center', out.width="80%", warning=F, message=F, echo=F}
ggplot(height_data, aes(y = height)) + 
  geom_boxplot(aes(fill = country), col = "black") +
  theme_minimal(base_size = 15) 
```



### Example, continued

Let R do the work for you: 

```{r}
t.test(height_data_wide %>% pull(usa), 
       height_data_wide %>% pull(dutch), 
       alternative = "two.sided")
```

### Example, continued
Note that `t.test` gives you both the t-test results (t-statistic (called "t" in 
the output), df, and p-value), as well as the 95% CI. We got both because we 
performed a two-sided test.


### Example: Transgenic chickens

Infection of chickens with the avian flu is a threat to both poultry production
and human health. A research team created transgenic chickens resistant to avian
flu infection. Could the modification affect the chicken in other ways? The 
researchers compared the hatching weights (in grams) of 45 transgenic chickens
and 54 independently selected commercial chickens of the same breed.

```{r, echo=FALSE}
transgenic <- c(38.8, 39.0, 39.7, 40.0, 40.8, 40.9, 41.0, 41.0, 41.0, 42.5, 42.6, 43.0,
                43.0, 43.4, 43.5, 43.5, 43.8, 44.4, 44.7, 44.7, 44.7, 45.3, 45.7, 45.8, 
                46.4, 46.5, 46.6, 46.7, 46.7, 46.8, 46.9, 47.1, 47.1, 47.1, 47.3, 47.6,
                47.7, 48.1, 48.3, 49.3, 49.3, 49.8, 50.3, 50.9, 52.1)

commercial <- c(36.7, 37.1, 38.9, 39.5, 39.5, 39.8, 40.0, 40.2, 40.3, 40.5, 40.5, 40.7,
                41.1, 41.2, 41.5, 41.5, 41.6, 41.6, 41.7, 42.4, 43.1, 43.3, 43.3, 43.4,
                43.7, 44.1, 44.2, 45.2, 45.3, 45.4, 46.0, 46.1, 46.4, 46.6, 46.6, 46.9, 
                47.3, 47.5, 48.1, 48.2, 48.4, 48.6, 49.0, 49.1, 49.3, 49.6, 50.1, 50.2,  
                50.4, 50.6, 52.2, 53.0, 55.5, 56.4)

chicken_data <- data.frame(weight = c(transgenic, commercial), 
                           type = c(rep("transgenic", 45), rep("commercial", 54)))
```

### Example: Transgenic chickens
```{r, fig.align='center', out.width="80%", echo=F}
ggplot(chicken_data, aes(y = weight)) + geom_boxplot(aes(fill = type)) + theme_minimal(base_size = 15)

ggplot(chicken_data, aes(x = weight)) + geom_histogram(aes(fill = type), binwidth = 3, col = "black") +
  facet_wrap(~type, nrow = 2) + theme_minimal(base_size = 15)
```

### Estimate the size of the difference between the two means

```{r}
means <- chicken_data %>% 
  group_by(type) %>% 
  summarise(mean_weight = mean(weight))

diff_means <- means[1, 2] - means[2, 2]
diff_means
```

The estimated mean difference is -0.153 grams.

### Estimate the standard error

$SE = \sqrt{\frac{s_1^2}{n_1} + \frac{s_2^2}{n_2}}$

```{r}
chicken_stats <- chicken_data %>% 
  group_by(type) %>% 
  summarise(mean_weight = mean(weight),
            sd_weight = sd(weight), 
            n = length(weight)) 
```
 Use the output to calculate the SE:
 
$SE = \sqrt{\frac{4.568872^2}{54} + \frac{3.320836^2}{45}} = 0.7947528$

### Calculate the t-statistic

$$t=\frac{(\bar{x}_1 - \bar{x}_2)-(\mu_1 - \mu_2)}{\sqrt{\frac{s_1^2}{n_1} + \frac{s_2^2}{n_2}}}$$
$$t=\frac{(44.98889 - 45.14222)-(0)}{ 0.7947528} = -0.1929279$$
What is the chance of observing the t-statistic -0.193 on the t-distribution 
with the appropriate degrees of freedom?

To answer this, we would need to calculate the degrees of freedom using that 
crazy formula. We won't do this. Instead, we will ask R to do the test for us 
(and verify that our calculated t-statistic matches R's test)

### t.test in R

Pay attention to the arguments specified by `t.test`. The first argument is the 
weight data for the commercial chickens and the second argument is the weight 
data for the transgenic chickens.

```{r, echo=FALSE}
commercial_weight <- chicken_data %>% filter(type == "commercial") %>% pull(weight)
transgenic_weight <- chicken_data %>% filter(type == "transgenic") %>% pull(weight)
```

```{r}
t.test(commercial_weight, transgenic_weight, alternative = "two.sided")
```

### Two sample t-test 
To illustrate this example, I have simulated data for males and females using the mean and standard deviation of weights in the United States taken from the CDC NHANES data

```{r indwt, echo=FALSE}
# Read CSV into R
weights <- read.csv(file="weights.csv", header=TRUE, sep=",")
```
```{r, echo=TRUE}
weights %>% group_by(sex) %>% summarise(sample_mean = mean(weight1), 
                                                sample_sd = sd(weight1), 
                                                length = length(weight1))
```

### Two sample t-test
I can overlay the histograms for these data with this code:

ggplot(weights,aes(x=weight1)) + 

    geom_histogram(data=subset(weights,sex == 'M'),binwidth=5,fill="dark green", col="green") +

    geom_histogram(data=subset(weights,sex == 'F'),binwidth=5,fill = "blue", col="black") +

  theme_minimal(base_size = 15)

Notice that I am using two geom_histogram statements to lay the histograms on top of one another rather than using a "fill" statement in one geom_histogram.  

### Two sample t-test  
```{r ingraph,echo=FALSE, out.width="80%"}
ggplot(weights,aes(x=weight1)) + 
    geom_histogram(data=subset(weights,sex == 'M'),binwidth=5,fill="dark green", col="green") +
    geom_histogram(data=subset(weights,sex == 'F'),binwidth=5,fill = "blue", col="black") +
  theme_minimal(base_size = 15)
```

### Two sample t-test 
And a Student's T test will show that this difference is statistically significant - notice the syntax here
```{r indwt2, echo=TRUE}
t.test(weights$weight1~weights$sex, alternative="two.sided")
```


### Independent vs. non-independent samples
Using r to graph the post-trial weight against the pre-trial weight we can see that these are correlated

```{r wtscatter, echo=FALSE, out.width="80%"}
plot(weights$weight1,weights$weight2)
```


### Independent vs. non-independent samples
```{r indwt3, echo=TRUE}
t.test(weights$weight1, weights$weight3, data=weights)
```


### Paired t-test
```{r, echo=TRUE}
weights %>% summarise(dif_mean = mean(dif2), dif_sd = sd(dif2),
                      wt1_mean=mean(weight1),wt1_sd=sd(weight1),
                      wt3_mean=mean(weight3),wt3_sd=sd(weight3))
```

### Paired t-test
Notice the syntax here:
```{r indwt4, echo=TRUE}
t.test(weights$weight1, weights$weight3,data=weights, paired=TRUE)
```

### Distribution of differences

```{r ingraph2,echo=FALSE, out.width="80%"}
ggplot() +
geom_histogram(data = weights, aes(x = weight3), binwidth=5, fill="green") +
geom_histogram(data = weights, aes(x = dif2), binwidth=5, fill="blue") +
 labs( x = "Weight in kg") +
  theme_minimal(base_size = 15)
```



### Cholestorol measurements following two alternative diets - 

Suppose you received the following graphic illustrating cholesterol measurements following two alternate diets. What do you think about these data? 

```{r make-dataset, echo=F, message=F, warning=F, out.width="80%"}
# Eleven study participants were randomized to each diet, and you're tasked with 
# figuring out whether there is a difference between the diet's average cholesterol levels.

library(tidyverse)
library(RColorBrewer)

set.seed(123)
trt_a <- c(155, 180, 190, 192, 203, 201, 207, 208, 217, 228, 237)

# Model parameters: You can play around with these to see how they affect your analysis.
signal <- 7 # note: this is the true underlying mean difference we will estimate
noise <- 4  # note: 4/sqrt(n) is the true SE of the mean = 1.21, where n = length(trt_a) = 11

trt_b <- trt_a + rnorm(length(trt_a), signal, noise)

chol_dat <- data.frame("A" = trt_a, "B" = trt_b, "id" = 1:length(trt_a))

chol_dat_long <- chol_dat %>% gather(A, B, key = "diet", value = "cholesterol")
```

```{r, out.width="60%", echo=F, fig.align='center'}
ggplot(chol_dat_long, aes(diet, cholesterol)) + 
  geom_jitter(width = 0.03, pch = 21, size = 5, fill = "#2ca25f", alpha = 0.8) + 
  labs(y = "Cholesterol", x = "Diet") +
  theme_minimal(base_size = 15)
```

### Cholestorol measurements following two alternative diets - 

```{r, out.width="60%", echo=F, fig.align='center'}
means <- chol_dat_long %>% 
  group_by(diet) %>% 
  summarise(mean = round(mean(cholesterol), 1),
            median = round(median(cholesterol), 1))

ggplot(chol_dat_long, aes(diet, cholesterol)) + 
  geom_boxplot(fill = "transparent") +
  geom_jitter(width = 0.03, pch = 21, size = 5, fill = "#2ca25f", alpha = 0.8) + 
  labs(y = "Cholesterol", x = "Diet") +
  geom_text(data = means, aes(x = diet, y = 260, label = paste0("Mean: ", mean)), size = 5) +
  theme_minimal(base_size = 15)
```

- What do you notice about the variability between participants under each diet?
- What is the mean difference? 

```{r, echo=F, eval=F}
#There is a lot of between subject variability. This variation has nothing to do with the treatments.
#The mean difference is 7.8 units.
```

### Cholestorol measurements following two alternative diets - 

An independent t-test reveals no evidence against the null hypothesis of no difference between the diets:

```{r, out.width="80%", echo=F, fig.align='center'}
indep_t <- t.test(x = chol_dat %>% pull(A), 
                       y = chol_dat %>% pull(B), mu = 0, 
                       alt = "two.sided", paired = F, 
                       conf.level = 0.95)

indep_t
```

### Better visualization for a very small study

Now, what do you notice about the paired data?

```{r, out.width="80%", echo=F, fig.align='center'}
chol_dat_long <- chol_dat_long %>% mutate(diet_n = ifelse(diet == "A", 1, 2))
means <- means %>% mutate(diet_n = ifelse(diet == "A", 1, 2))

ggplot(chol_dat_long, aes(diet_n, cholesterol)) + 
  geom_line(aes(group = id), lty = 3, lwd = 1) +
  geom_point(aes(fill = as.factor(id)), pch = 21, size = 5, alpha = 0.8) + 
  scale_fill_brewer(palette = "Spectral") +
  labs(y = "Cholesterol", x = "Diet") +
  geom_text(data = means, aes(x = diet_n, y = 260, label = paste0("Mean: ", mean)), size = 5) +
  theme_minimal() +
  guides(fill = guide_legend("ID")) +
  scale_x_continuous(breaks = c(1, 2), labels = c("A", "B"), limits = c(0.5, 2.5)) +
  theme_minimal(base_size = 15)
  theme(panel.grid.minor.x = element_blank())
```

```{r, echo=F, eval=F}
# There is a consistent increase in cholesterol under Diet B. 
# We also know that the mean of the individual differences is equal to the difference
# of the means --> the average of the individual differences is also 209.4-201.6 = 7.8
# So what is different about the test? It's the standard error. 
```

### apply a paired t-test


- The observed value of the test statistic is: 
$t = \frac{\bar{x}_d-0}{s_d/\sqrt{n}}$

- It can be compared to a critical value from the t distribution with $n-1$ degrees of freedom

### Calculate the test statistic, p-value, and 95% confidence interval {.demphasize}

- First let's have a look at the dataset as is:

```{r peak-data} 
head(chol_dat)
```

### Calculate the test statistic, p-value, and 95% confidence interval {.demphasize}

- We can use functions from the library `dplyr` to calculate the test statistic
- Use `mutate` to calculate each participant's difference:

```{r mutate-diff, echo=TRUE}
chol_dat <- chol_dat %>%mutate(diff = B - A)
head(chol_dat)
```

### Calculate the test statistic, p-value, and 95% confidence interval

- Then use `summarize` to calculate the mean difference ($\hat{\mu}_d$), its
standard error ($\hat{s}_d/\sqrt{n}$), and the observed t-statistic:

```{r calc-t-stat, echo=TRUE}
summary_stats <- chol_dat %>% 
  summarize(mean_diff = mean(diff),  # mean difference
            std_err_diff = sd(diff)/sqrt(n()),  # SE of the mean 
            t_stat = mean_diff/std_err_diff)   # test statistic
summary_stats
```

### Calculate the test statistic, p-value, and 95% confidence interval {.tiny}

What is the <b class = "greentext">probability</b> of observing a t-stat $\geq$ `r round(summary_stats$t_stat, 2)` or $\leq$ `r round(-summary_stats$t_stat, 2)` using the `pt` command.

```{r graph-t, out.width="80%", echo=F, fig.align='center'}
##You do not need to know how to do this
ggplot(data.frame(x = rt(100000, 10)), aes(x = x)) + 
  stat_function(fun = dt, args = list(df = 10)) + 
  scale_x_continuous(limits = c(-9, 9), 
                     breaks = c(-8, -4, 0, 4, 8), 
                     labels = c(-8, -4, 0, 4, 8))  +
  geom_vline(xintercept = c(-summary_stats$t_stat, summary_stats$t_stat), lty = 3) +
  geom_area(stat = "function", fun = dt, args = list(df = 10), fill = "blue", xlim = c(-5, -summary_stats$t_stat)) +
  geom_area(stat = "function", fun = dt, args = list(df = 10), fill = "blue", xlim = c(summary_stats$t_stat, 5)) + 
  labs(x = "t distribution") + 
  theme(panel.grid.minor = element_blank()) 
```

```{r calc-pval}
pt(q = summary_stats %>% pull(t_stat), lower.tail = F, df = 10) * 2
```

### Calculate the test statistic, p-value, and 95% confidence interval {.demphasized}

- To calculate the 95% confidence interval, we need to know the quantile of the 
t distribution such that 2.5% of the data lies above or below it.

- Ask R: What is the <b class = "greentext">quantile</b> such that 97.5% of the t-distribution is below it on 10 degrees of freedom using the `qt` command.

### Calculate the test statistic, p-value, and 95% confidence interval {.demphasized}
```{r calc-CI, echo=TRUE}
q <- qt(p = 0.975, lower.tail = T, df = 10)
q
ucl <- summary_stats %>% pull(mean_diff) + (q * summary_stats %>% pull(std_err_diff))   
lcl <- summary_stats %>% pull(mean_diff) - (q * summary_stats %>% pull(std_err_diff))  
c(lcl, ucl)
```

The confidence interval is (`r lcl`, `r ucl`).

### Calculate the test statistic, p-value, and 95% confidence interval {.demphasized}

- Or, have R do the work for you! Just be sure to specify that `paired = T`.
```{r lazy-way}
paired_t <- t.test(chol_dat %>% pull(B), chol_dat %>% pull(A), 
                   alternative = "two.sided", mu = 0, paired = T)
```

### Calculate the test statistic, p-value, and 95% confidence interval {.demphasized}
```{r lazy2}
paired_t
```


### coding notes

A one sample t- test will take the form:

t.test(x = **x variable**, alternative = **greater, less or two.sided**, mu = **null hypothsis value**)

A two sample t-test will take the form:

t.test(**first sample data**, **second sample data**, alternative = **greater, less or two.sided**)

A paired t-test will take the form:

t.test(**first data points**, **second datapoints**, alternative = **greater, less or two.sided**, paired=TRUE)


