---
title: "3+ Sample testing with continuous outcomes:  ANOVA"
---

<!-- libraries -->
```{r,include=FALSE,purl=FALSE}
library(knitr) # for include_graphics() 
library(dplyr)
library(forcats)
```

### A) Is there a difference between these means?

Describe why you do or do not think so.

```{r, out.width="80%", echo=FALSE, warning=F, message=F}
library(ggplot2)
library(tidyr)
library(dplyr)
set.seed(1)

group_1 <- rnorm(n = 10, mean = 15, sd = 3)
group_2 <- rnorm(n = 10, mean = 15, sd = 3)
group_3 <- rnorm(n = 10, mean = 15, sd = 3)

no_diff <- data.frame(group_1, group_2, group_3)

no_diff_narrow <- no_diff %>% gather(key = "Group", value = "Measure", group_1:group_3)

means <- no_diff_narrow %>% group_by(Group) %>% summarise(Measure = mean(Measure))

ggplot(no_diff_narrow, aes(x = Group, y = Measure)) + geom_point(aes(pch = "data")) + 
  geom_point(data = means, col = "red", size = 2, aes(pch = "mean estimate")) + theme_minimal(base_size = 15) + 
  scale_shape_manual(values = c(4, 20)) +
  geom_text(data = means, col = "red", aes(label = paste("mean estimate =\n", round(Measure, 2))), nudge_x = 0.25) +
  scale_y_continuous(limits = c(8, 40))
```

### B) Is there a difference between these means?

Describe why you do or do not think so.


```{r, out.width="80%", echo=FALSE, warning=F, message=F}
group_1 <- rnorm(n = 10, mean = 15, sd = 3)
group_2 <- rnorm(n = 10, mean = 20, sd = 3)
group_3 <- rnorm(n = 10, mean = 27, sd = 3)

diff <- data.frame(group_1, group_2, group_3)

diff_narrow <- diff %>% gather(key = "Group", value = "Measure", group_1:group_3)

means_2 <- diff_narrow %>% group_by(Group) %>% summarise(Measure = mean(Measure))

ggplot(diff_narrow, aes(x = Group, y = Measure)) + geom_point(aes(pch = "data")) + 
  geom_point(data = means_2, col = "red", size = 2, aes(pch = "mean estimate")) + theme_minimal(base_size = 15) + 
  scale_shape_manual(values = c(4, 20)) +
  geom_text(data = means_2, col = "red", aes(label = paste("mean estimate =\n", round(Measure, 2))), nudge_x = 0.25) +
  scale_y_continuous(limits = c(8, 40))
```

### C) Is there a difference between these means?

Describe why you do or do not think so.

```{r, out.width="80%", echo=FALSE, warning=F, message=F}
group_1 <- rnorm(n = 10, mean = 15, sd = 3)
group_2 <- rnorm(n = 10, mean = 20, sd = 3)
group_3 <- rnorm(n = 10, mean = 15, sd = 3)

diff_2 <- data.frame(group_1, group_2, group_3)

diff_2_narrow <- diff_2 %>% gather(key = "Group", value = "Measure", group_1:group_3)

means_3 <- diff_2_narrow %>% group_by(Group) %>% summarise(Measure = mean(Measure))

ggplot(diff_2_narrow, aes(x = Group, y = Measure)) + geom_point(aes(pch = "data")) + 
  geom_point(data = means_3, col = "red", size = 2, aes(pch = "mean estimate")) + theme_minimal(base_size = 15) + 
  scale_shape_manual(values = c(4, 20)) +
  geom_text(data = means_3, col = "red", aes(label = paste("mean estimate =\n", round(Measure, 2))), nudge_x = 0.25) +
  scale_y_continuous(limits = c(8, 40))
```


### D) Is there a difference between these means?

Describe why you do or do not think so.

```{r, out.width="80%", echo=FALSE, warning=F, message=F}
group_1 <- rnorm(n = 10, mean = 15, sd = 7)
group_2 <- rnorm(n = 10, mean = 20, sd = 7)
group_3 <- rnorm(n = 10, mean = 27, sd = 7)

diff_3 <- data.frame(group_1, group_2, group_3)

diff_3_narrow <- diff_3 %>% gather(key = "Group", value = "Measure", group_1:group_3)

means_4 <- diff_3_narrow %>% group_by(Group) %>% summarise(Measure = mean(Measure))

ggplot(diff_3_narrow, aes(x = Group, y = Measure)) + geom_point(aes(pch = "data")) + 
  geom_point(data = means_4, col = "red", size = 2, aes(pch = "mean estimate")) + theme_minimal(base_size = 15) + 
  scale_shape_manual(values = c(4, 20)) +
  geom_text(data = means_4, col = "red", aes(label = paste("mean estimate =\n", round(Measure, 2))), nudge_x = 0.25) +
  scale_y_continuous(limits = c(8, 40))
```

## Descriptive plots


### Box plot

```{r, out.width="70%", fig.align='center', echo=FALSE}
ggplot(diff_3_narrow, aes(x = Group, y = Measure)) + 
  geom_boxplot() +
    geom_point() +
  theme_minimal(base_size = 15)
```

### Density plot

```{r, out.width="60%", fig.align='center', echo=FALSE}
ggplot(diff_3_narrow, aes(x = Measure)) + 
  geom_density(aes(fill = Group), alpha = 0.5) +
  theme_minimal(base_size = 15)
```

### Histograms with facet wrap
```{r, out.width="60%", fig.align='center', echo=FALSE}
ggplot(diff_3_narrow, aes(x = Measure)) + 
  geom_histogram(aes(fill = Group), col = "white", binwidth = 2.5) +
  theme_minimal(base_size = 15) + facet_wrap(~ Group, nrow = 3)
```


### ANOVA in R: use `aov()`, then `tidy()` it up!

- `aov()` stands for analysis of variance

The general syntax for the ANOVA is:

aov(outcomevariable ~ groupvaraible, data=dataset)

We will save the output of this as an object and then use tidy(object) to get the output we want.

reference: https://broom.tidyverse.org/reference/anova_tidiers.html



## Example:  Cannabis to treat brain cancer

### The data

```{r, echo=FALSE}
treatment <- c(rep("Irradiation", 4), rep("Cannabinoids", 5), rep("Both", 6),
               rep("Neither", 7))

tumor_volume <- c(30, 46, 46, 95, # Irradiation
                  12, 14, 16, 41, 47, # Cannabinoids 
                  5, 4, 4, 4, 10, 9, # Both
                  24, 30, 43, 51, 62, 32, 96) # Neither

cancer_data <- data.frame(treatment, tumor_volume)
```
```{r}
head(cancer_data)
```


```{r, out.width="80%", fig.align='center'}
# specify the order of the treatment groups for plotting
library(forcats) 
cancer_data <- cancer_data %>% 
  mutate(trt_order = fct_relevel(treatment, c("Neither", "Irradiation", 
                                              "Cannabinoids", "Both")))
```

### Look at summary statistics
```{r}
# calculate the means and SD for each group
summary_stats <- cancer_data %>% 
  group_by(trt_order) %>% 
  summarise(mean_vol = mean(tumor_volume),
            sd_vol = sd(tumor_volume), 
            samp_size = n())

summary_stats
```


### Graph the data
```{r, out.width="80%", fig.align='center', echo=FALSE}
ggplot(cancer_data, aes(x = trt_order, y = tumor_volume)) + 
  geom_jitter(pch = 4,width = 0.1) + 
  geom_point(data = summary_stats, aes(y = mean_vol, col = "Mean"), pch = 19) +
  labs(y = "Tumor volume (mm cubed)", x = "Treatment") +
  theme_minimal(base_size = 15)
```

### ANOVA in R: use `aov()`, then `tidy()` it up!
```{r}
library(broom)
cancer_anova <- aov(formula = tumor_volume ~ treatment, data = cancer_data)
tidy(cancer_anova)
```

### P value of the F test 
- The p-value of the ANOVA F statistic is always the area to the right of the 
test statistic

```{r}
pf(6.699489, df1 = 3, df2 = 22 - 4, lower.tail = F)
```


## After a significant ANOVA...

###  Bonferroni for our example
So we would conduct each pairwise test:
```{r}
both<-cancer_data%>%filter(treatment=="Both")%>%select(tumor_volume)
neither<-cancer_data%>%filter(treatment=="Neither")%>%select(tumor_volume)
can<-cancer_data%>%filter(treatment=="Cannabinoids")%>%select(tumor_volume)
irr<-cancer_data%>%filter(treatment=="Irradiation")%>%select(tumor_volume)
n_b<-t.test(neither, both, alternative="two.sided")
c_b<-t.test(can,both, alternative="two.sided")
i_b<-t.test(irr, both, alternative="two.sided")
n_c<-t.test(neither, can, alternative="two.sided")
n_i<-t.test(neither, irr, alternative="two.sided")
c_i<-t.test(can, irr, alternative="two.sided")
```

###  Bonferroni for our example
and compare the p-value of this test to an alpha of 0.0083
```{r}
t_statistics<-c(n_b$statistic, c_b$statistic, i_b$statistic, n_c$statistic, n_i$statistic, c_i$statistic)
p_values<-c(n_b$p.value, c_b$p.value, i_b$p.value, n_c$p.value, n_i$p.value)
t_statistics
p_values
```

## Tukey's HSD

### `TukeyHSD()` to calculate the differences in R

You can think of the TukeyHSD() as a wrap-around for the anova, you can either nest the statements like this:

TukeyHSD(aov(outcome ~ group))

or save the ANOVA as an object and use that in the statement:

modelresult<-aov(outcome ~ group)

TukeyHSD(modelresult, overall_alpha)



### `TukeyHSD()` to calculate the differences in R
Here is the R code and output for the cancer example:

```{r, out.width="80%", fig.align='center'}
diffs <- TukeyHSD(cancer_anova, conf.level = 0.05) %>% tidy()
diffs%>%select(contrast, adj.p.value)
```


### Visualize the pairwise differences

```{r, fig.align='center', out.width="90%", echo=F}
#Students do not need to know how to do this
ggplot(diffs, aes(x = contrast, y = estimate)) + geom_point() +
  geom_segment(aes(y = conf.low, yend = conf.high, xend = contrast)) +
  theme_minimal(base_size = 15) +
  geom_hline(aes(yintercept = 0), lty =2) +
  geom_text(aes(label = paste0("p-value:\n ", round(adj.p.value, 2))), nudge_x = 0.3) +
  labs(y = "Estimated difference", x = "") + 
  scale_x_discrete(labels = c("Cannabinoids\n vs. both", "Irradiation\n vs. both",
                              "Irradiation vs.\ncannabinoids", "Neither\n vs. both", 
                              "Neither vs.\n Cannabinoids", "Neither vs.\n Irradiation"))
```

### Review raw data for comparison
```{r, out.width="80%", fig.align='center', echo=FALSE}
ggplot(cancer_data, aes(x = trt_order, y = tumor_volume)) + 
  geom_jitter(pch = 4,width = 0.1) + 
  geom_point(data = summary_stats, aes(y = mean_vol, col = "Mean"), pch = 19) +
  labs(y = "Tumor volume (mm cubed)", x = "Treatment") +
  theme_minimal(base_size = 15)
```

