---
title: "Non-parametrics"
---

<!-- libraries -->
```{r,include=FALSE,purl=FALSE, warning=FALSE, message=FALSE}
library(knitr) # for include_graphics() 
library(dplyr)
library(forcats)
library(ggplot2)
```


<!-- libraries -->
```{r,include=FALSE,purl=FALSE}
library(knitr) # for include_graphics() 
library(dplyr)
library(forcats)
library(readr)
library(ggplot2)
library(broom) 
library(tidyr)
library(ggplot2)
library(tibble)
```


### Wilcoxon Rank-Sum example :phenylketonuria 
Normalized mental age scores for children with phenylketonuria

Group 1:  "low exposure" < 10.0 mg/dl

Group 2:  "high exposure" >= 10.0 mg/dl


### Wilcoxon Rank-Sum :phenylketonuria 
```{r read pku, echo=FALSE}
nMA<- c(34.5, 37.5, 39.5, 40,45.5, 47,47,47.5,48.7, 49, 51,51,52,53,54,54,55,56.5,57,58.5,58.5,28,35,37,37,43.5,44,45.5,46,48,48.3,48.7,51,52,53,53,54,54,55)
Group<-c((rep("low",21)),rep("high",18))

pku <- data.frame(Group,nMA)
head(pku)
```


### Wilcoxon Rank-Sum :phenylketonuria 
In this example there 18 High and 21 Low exposure individuals.
```{r}
group_by(pku,Group) %>%
  summarise(
    count = n(),
    median = median(nMA, na.rm = TRUE),
    IQR = IQR(nMA, na.rm = TRUE)
  )
```

### Wilcoxon Rank-Sum: PKU
If we graph the distributions with a density plot what do we notice?
```{r, echo=TRUE, out.width="60%", fig.align='center'}
ggplot(pku, aes(x = nMA)) + 
  geom_density(aes(fill = Group), alpha = 0.5) +
  theme_minimal(base_size = 15)
```

### Wilcoxon Rank-Sum: PKU
```{r, echo=TRUE, message=F, warning=FALSE}
wilcox.test(nMA ~ Group, data=pku,paired=F)
```

### Wilcoxon Rank-Sum vs T : NHANES example
Here I will again use the NHANES data as an example, looking at height by gender
```{r read, echo=TRUE}
# Read CSV into R
nhanes <- read.csv(file="nhanes.csv", header=TRUE, sep=",")
names(nhanes)
```


### Wilcoxon Rank-Sum vs T : NHANES example
```{r wil1, echo=FALSE}
boxplot(bmxht~gender,data=nhanes, main="Height in cm by gender")
malesht <- nhanes$bmxht[nhanes$gender=="Male"]
femalesht <- nhanes$bmxht[nhanes$gender=="Female"]
```

### Wilcoxon Rank-Sum vs T
```{r, echo=TRUE, out.width="60%", fig.align='center', message=FALSE, warning=FALSE}
ggplot(nhanes, aes(x = bmxht)) + 
  geom_density(aes(fill=gender), alpha=0.1) +
  theme_minimal(base_size = 15)
```

### Wilcoxon Rank-Sum vs T
```{r wil2, echo=TRUE}
t.test(malesht, femalesht, paired=F)
```


### Wilcoxon Rank-Sum vs T
```{r wilvst, echo=TRUE}
wilcox.test(malesht,femalesht,paired=F)
```


### Wilcoxon Sign rank
Steps:

1) Calculate the difference between each pair of observations 

2) Rank the difference by absolute value from smallest to largest (again, tie values get the average of the ranks).  Any pair where difference = 0 is thrown out.

3) Assign a "sign" for whether the difference was positive or negative

4) Take the sum of the positive ranks and the sum of the negative ranks (the smaller sum is denoted with a T).  


### Wilcoxon Sign rank
Under the null hypothesis that the difference is 0, we would expect the sample to have equal numbers of positive and negative ranks with equivalent sums.  This expectation is tested against the statistic
$$Z_{T}=\frac{T-\mu_{T}}{\sigma_{T}}$$

Where
$$\mu_{T}=\frac{n(n+1)}{4}$$
and 
$$\sigma_{T}=\sqrt{\frac{n(n+1)(2n+1)}{24}}$$


### Wilcoxon Sign rank:  Example Pre and post test
```{r sign, echo=FALSE}
Change<-c(12,13,-2,-1,10,-3,-1,13,-1,-2,20,28,0)
test1<-c(65,87,77,90,70,84,92,83,85,91,68,72,81)
test2<-c(77,100,75,89,80,81,91,96,84,89,88,100,81)
prepost<-data.frame(Change,test1,test2)
```

|Time 1 |Time 2|
|---------|------|
|65       |77   |
|87       |100     |
|77        |75    |
|90       |89      |
|70       |80    |
|84       |81    |
|92       |91    |
|83       |96    |
|85       |84    |
|91       |89    |
|68       |88    |
|72       |100    |
|81       |81    |
|---------|------|

### Sign rank example
```{r, echo=FALSE, out.width="60%", fig.align='center', message=FALSE, warning=FALSE}
ggplot(prepost) + 
  geom_density(aes(test1), col="blue") +
  geom_density(aes(test2), col="green") +
  theme_minimal(base_size = 15)
```
### Sign Rank example: calculate difference and sign


|Time 1 |Time 2| Difference | sign |
|---------|------|----------|------|
|65       |77    |12  |+|
|87       |100   |13  |+|
|77        |75   |-2  |-|
|90       |89    |-1  |-|
|70       |80    |10  |+|
|84       |81    |-3  |-|
|92       |91    |-1  |-|
|83       |96    |13  |+|
|85       |84    |-1  |-|
|91       |89    |-2  |-|
|68       |88    |20  |+|
|72       |100   |18  |+|
|81       |81    |0   |?|
|---------|------|----|------|

### Sign Rank example: sort by absolute value and assign rank

|Time 1 |Time 2| Difference | sign | rank|
|---------|------|----------|------|-----|
|90       |89   |-1  |-|2|
|92       |91   |-1  |-|2|
|85       |84   |-1  |-|2|
|77       |75    |-2  |-|4.5|
|91       |89    |-2  |-|4.5|
|84       |81    |-3  |-|6|
|70       |80    |10  |+|7|
|65       |77    |12  |+|8|
|87       |100  |13  |+|9.5|
|83       |96   |13  |+|9.5|
|72       |100    |18  |+|11|
|68       |88   |20  |+|12|
|81       |81    |0   |?|**drop**|
|---------|------|----|------|
 
### Sign Rank example: sum the positive and negative ranks
Negative signs 

|Time 1 |Time 2| Difference | sign | rank|
|---------|------|----------|------|-----|
|90       |89   |-1  |-|2|
|92       |91   |-1  |-|2|
|85       |84   |-1  |-|2|
|77       |75    |-2  |-|4.5|
|91       |89    |-2  |-|4.5|
|84       |81    |-3  |-|6|
|---------|------|----|------|

Sum of Negative sign ranks is 21

### Sign Rank example: sum the positive and negative ranks

|Time 1 |Time 2| Difference | sign | rank|
|---------|------|----------|------|-----|
|70       |80    |10  |+|7|
|65       |77    |12  |+|8|
|87       |100  |13  |+|9.5|
|83       |96   |13  |+|9.5|
|72       |100    |18  |+|11|
|68       |88   |20  |+|12|
|---------|------|----|------|

Sum of the positive sign ranks is 57


### Wilcoxon Sign rank:  Example
Our expectation would be 
$$\mu_{T}=\frac{n(n+1)}{4}=\frac{12(12+1)}{4}=39$$
**remember that we had 13 observations, but we dropped one because the scores at times 1 and 2 were the same**
and 
$$\sigma_{T}=\sqrt{\frac{n(n+1)(2n+1)}{24}}=\sqrt{\frac{12(12+1)(2*12+1)}{24}}=12.75$$

### Wilcoxon Sign rank:  Example
And we compare our expectation to the smaller rank value (Sum of negative ranks was 21, sum of positive ranks was 57)
$$Z_{T}=\frac{T-\mu_{T}}{\sigma_{T}}=\frac{21-39}{12.75}=-1.412$$

```{r, echo=TRUE}
2*pnorm(-1.412)
```

### Wilcoxon Sign-rank in R

The general syntax will be:

wilcox.test(group1, group2, paired=T)

or 

wilcox.test(outcome ~ group, paired=T)


### Wilcoxon Sign rank:  Example

```{r, sign2, echo=TRUE}
wilcox.test(test1,test2,paired=T, correct=FALSE)
```

### Wilcox Sign rank: compare to T
```{r spairt,echo=TRUE}
t.test(test1,test2,paired=TRUE)
```

### Wilcox Sign rank:  Compare to T
With this study, our sample is 13 and the distribution of changes looks like this - remember that the 0 difference value gets thrown out of sign rank test:
```{r wtot, echo=TRUE, out.width="80%", fig.align='center'}
hist(Change)
```

### Kruskal Wallis
This test can be thought of as an extension of the rank sum test as it is based on the Rank-sum test.
We will not do this one by hand.  
In R the syntax is generally:

kruskal.test(outcome ~ group, dataset)

### Kruskal Wallis
```{r, echo=FALSE}
##students you could tinker with this code to see what happens if you add outliers, or shift the outcomes for a group
treatment <- c(rep("Group1", 4), rep("Group2", 5), rep("Group3", 6), rep("Group4", 7))
outcome <- c(30, 46, 99, 95,  # group1
                  3, 14, 16, 41, 47, #group2
                  5, 4, 4, 4, 10, 9, #group3
                  24, 30, 43, 51, 62, 32, 96 ) #group4 

KW_data <- data.frame(treatment, outcome)
kruskal.test(outcome~ treatment, data=KW_data)
```


